Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-24 manuscript/response_to_reviews.Rmd

We have revised our manuscript, “Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest” (previously “Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest”), according to the recommendations of reviewers. This has resulted in an easier to read manuscript with more robust support.

We addressed all of the reviewers’ points, as detailed below, resulting in several major changes:

* At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years). Results were similar, and conclusions the same. We now present those results as a secondary model (mostly in SI).

* We dropped crown position as a factor in the statistical models of drought resistance, and instead include it as part of an enhanced focus on the differences in microenviroment experienced by trees of different height (now in Table 1; Fig. 2). This change was prompted by comments by two reviewers (R1, R3) that canopy position could not be reliably projected back to 1966, and our own conclusion that it could not be reliably disentangled from height.

@@

* We shifted to more figures and fewer tables in the main article (as requested by R1, R3, and R4). We added two figures to the main text: one showing Rt differences across species (Fig. 3) and one visualizing the model results (Fig. 4). We moved former tables 4 and 5 to the SI. We also added several new figures and tables to the SI to provide information requested by the reviewers.

We hope that this manuscript will now be found suitable for publication in *New Phytologist*, and we look forward to your response.

@@

Comments to the Author

The study addresses the relationship between drought resistance (Rt) calculated from radial growth, and different traits and microenvironmental characteristics on 12 tree species from one forest in the Eastern US. Both the dataset and the study are meaningful, and the relationship found between Rt and height of biological importance. However, I think there are a number of issues that need to be worked out.

It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought. This metric has been much used lately as one of the so-called resilience indices (I wonder why other Rs indices where not included in the manuscript). In the paper, the authors use Rt similarly to most studies in the literature, which is fine. Yet, Rt (and particularly the other resilience indices based on radial growth) does not only depend on the individual response to drought events, but also on stand dynamics around target trees. For this reason, if stands around trees are not comparable or show changes after some disturbance (e.g. the drought events analysed) differently, this may bias comparisons of Resilience indices (although it is true that Rt should be less influenced than the other indices). The problem is that there is not such stand competition data for old droughts. This does not invalidate the study but I think the authors should discuss the issue, particularly for the analysis and comparisons of single droughts. I would suggest including also other Resilience indices, not only Rt, to enrich the discussion.

**To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA). Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4, and in the revised manuscript (Fig. S5). In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric. While this unquestionably leads to biased estimates for some trees, we would not expect it to bias our analysis; rather, it is one of the reasons that the amount of variation that can be explained by analyses such as these is relatively low. We now include the following content on ARIMA in the manuscript:**

@@

**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). The two metrics were strongly correlated (Fig. **S5**). Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info."**

**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance. While it would be interesting to examine resilience, this would be beyond the scope of the current paper. (2) @desoto_low_2020 used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought. In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought. In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought. These findings would argue that resistance is a more important variable for angiosperms. (3) As the reviewer notes, resilience indices would probably tend to be more sensitive to a tree's neighborhood and prior conditions. We now state the logic for focusing on $Rt$ in the methods section:**

@@

**We have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **

Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included. And I suggest discussing only the model with all years together. This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation. This would help to ensure a better use of the Resilience indices.

**In the spirit of this comment (simplifying presentation), we have created a new figure (*Fig. 4*) visualizing the results of the best model. We moved Table 5 to the Appendix (now Table S6, plus a parallel Table S7 for $Rt_{ARIMA}$).**

**We have also decreased the emphasis on differences across drought years, as follows: **

@@

**(1) Table 1 now shows just the results from the model with all drought years combined.**

**(2) To make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before). This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$). (Removed was xylem porosity, which had opposite effects in different droughts.)**

@@

**(3) The results and discussion now focus more on the results from the three drought years combined.**

**We do still present results from the three droughts separately (e.g., Figs. 1,3,4; Tables S4-S7), as we do feel that this makes the story richer.**

The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis. If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.

**We have moved this table to the SI. We modified the analysis approach such that separate models for individual variable tests (as previously shown in this table) are now limited to the species traits. While this table is (and was) the only display item (besides Table 1) giving these results, the results are either mostly negative (wood density, SLA, xylem porosity) or preliminary tests qualifying variables for inclusion in full model ($PLA_{dry}$ and $\pi_{tlp}$).**

@@

- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This 
should be rather straightforward and the authors would increase accuracy across species in height estimation.

**We have specified how many species used the interspecific allometry (n=2, JUNI and FRAM). Because of COVID-19, it would not be possible to go collect more data for these species.**

- Lines 166-167: please explain further the hypothesis and why.

**Crown position, and this hypothesis, have been removed from the analysis of $Rt$.**

@@

- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.

**We no longer attempt to use crown position to describe variation in drought resistance. Instead, we now use it to characterize *contemporary* differences in microenvironmental conditions faced by trees of different heights.**

- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest. They are indeed interesting to characterise the forest, but they do not add to individual tree variability. You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.

**Unless the reviewers and editor strongly disagree, we would request to keep this figure in the main manuscript. We have revised the manuscript such that this figure is better integrated with the hypothesis testing (Table 1) and interpretation of results. We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing this depiction of height profiles for a temperate forest. (Such patterns are not shown in some relevant textbooks: Gordon Bonan's 2016 *Ecological Climatology* [@bonan_ecological_2015] and Norman & Campbell's *Environmental Biophysics* [@campbell_introduction_1998]. If we're missing studies that show such vertical profiles, we'd appreciate the references.)**

@@

- Line 348: non-drought.

**Fixed.**

- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.

**We did not intend the meaning here as vulnerability *to mortality* (assuming that's the objection here). We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought."**

@@

- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.

**This text was removed, and we now clearly distinguish between drought resistance and vulnerablity to mortality throughout the manuscript.**

@@

73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.

**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought. While tree size can correlate with the depth of water extraction [@brum_hydrological_2019], the linkage is not consistent. Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999]. Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitratarak_roots_2018]."**

73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.

**We have updated our text to read: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011]."**

@@

112 wasn’t clear where the “ring porous more drought resistant” hypothesis came from. Global data show that ring porous species are almost entirely restricted to the north temperate zone, especially deciduous broadleaf forests (Wheeler et al., 2007). They are virtually absent in drylands, especially warm tropical drylands. Species in drylands are almost overwhelmingly diffuse porous; the high wood density species in chaparrals and in some desert plants can be semi ring porous (Olson et al., in press).

**This is a great point. The hypothesis refers to temperate deciduous biome, which we now clarify in the sentence in question:**

**"Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015] and would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category."**

**We dug into the ring/diffuse porous distinction more, and ended up making a few substantive changes that relate to this theme: (1) We added a new figure showing $Rt$ across species (Fig. 3). This illustrates that the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum. (2) We modified the criteria of selecting candidate variables for the full models to include the criterion that directions of responses must be consistent across droughts for inclusion in the full model. This resulted in exclusion of xylem porosity (Table S4)--a change that is also warranted by the fact that we only have two diffuse-porous species. (3) We added the following content to the discussion: "Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$)...Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive. There are two potential explanations for this discrepancy. First, other traits can and do override the influence of xylem porosity on drought resistance. Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]." **

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
final check of reviewer responses (up to line 268 in .rmd file)

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.

>>>>

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.
